---
title: "Analysis of Serum Metabolomes in Dogs with Exocrine Pancreatic Insufficiency"
output: html_notebook
---


```{r setup, include=F}
knitr::opts_chunk$set(echo = F, message=FALSE, warning=FALSE)

library(knitr)
library(dplyr)
library(pheatmap)
library(fgsea)
library(factoextra)
library(EnhancedVolcano) 
library(rstatix)
library(tidyr)
library(corrr)
library(RColorBrewer)
library(viridis)
library(visdat)
library(purrr)
library(ggpubr)
library(qvalue)
library(FactoMineR)
library(reshape2)
library(ggforce)
library(reticulate)
library(arsenal)
library(gtools)
library(mixOmics)
library(multcomp)
library(ggplus)
library(Hmisc)

#set plotting colors for Healthy vs EPI
plot_cols <- c("royalblue", "orange")

```

# **LOAD DATA**

## **SAMPLE METADATA**

```{r}
baseline_clinicVar <- read.csv("k9EPI_baseline_clinicVars.csv", row.names = 1)

baseline_clinicVar <- baseline_clinicVar[, -1]

baseline_clinicVar$Group <- factor(baseline_clinicVar$Group, levels = c("Healthy", "EPI"))

#groups for the serummetabolome data
EPI_groups <- read.csv("K9EPI_groups.csv", row.names = 1)

EPI_groups$Group <- factor(EPI_groups$Group, levels = c("Healthy", "EPI"))


```

How many dogs are in each group for the whole investigation?

```{r}
table(baseline_clinicVar$Group)
```

## **SERUM METABOLOMES**

```{r, include=F}
metab <- read.csv("K9EPI_serumMetab.csv", row.names = 1)

```

How many metabolites are contained in these data?

```{r}
paste("The (untargeted) serum metabolite dataset contains", nrow(metab), "named biochemicals")
```

## **METABOLITE ANNOTATION**

```{r, include=F}
#load metabolite annotation 
annot <- read.csv("K9EPI_metabAnnotation.csv", row.names = 1)

annot$SUPER.PATHWAY <- as.factor(annot$SUPER.PATHWAY)

length(levels(annot$SUPER.PATHWAY))

annot$SUB.PATHWAY <- as.factor(annot$SUB.PATHWAY)

```

```{r}
paste("Serum metabolites are distributed among", length(levels(annot$SUPER.PATHWAY)), "super-pathways.", sep = " ")
```

```{r}
super <- data.frame(table(annot$SUPER.PATHWAY))

names(super) <- c("Super Pathway", "Count")

#super <- super %>% arrange(desc(Count))

ggplot(data = super, aes(x= reorder(`Super Pathway`, -Count), y = Count, fill = `Super Pathway`)) + 
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(legend.position="right") + 
  theme(legend.title=element_text(size=12), 
    legend.text=element_text(size=10)) +
  labs(x="Metabolic Super-Pathway") +
  theme(axis.text.x=element_blank()) +
  ggtitle("Distribution of Serum Metabolites in Metabolic Super-Pathways")
```

# **ANALYZING DEMOGRAPHIC AND CLINICAL VARIABLES**

## **Demographic Variables**

### **Age**

Assess distribution of "age" with histogram and Shapiro-Wilk test

```{r}
hist(baseline_clinicVar$Age, main = "Distribution of Age")
shapiro_test(baseline_clinicVar$Age)
```

Summary and inferential statistics for age, breed, and reproductive status:

```{r, results='asis'}

summary(tableby(Group ~ anova(Age) + Breed + Sex, data = baseline_clinicVar, 
                numeric.stats=c("mean", "sd", "median","min", "max")))
```

There are significant differences in age, breed, or reproductive status between EPI dogs and healthy controls.

### **Clinical Signs**

```{r, results='asis'}
summary(tableby(Group ~ Diarrhea + Weight + Vomiting + Appetite, 
                data = baseline_clinicVar[baseline_clinicVar$Group != "Healthy", ]))
```

## **Medication History**

```{r, results='asis'}
summary(tableby(Group ~ Metronidazole + Tylosin + Cobalamin_sup + Folate_sup + Antacids + Probiotics, 
                data = baseline_clinicVar[baseline_clinicVar$Group != "Healthy", ]))

```

### **Serum cTLI**

```{r}
hist(baseline_clinicVar$Serum_cTLI, main = "Distrubition of Serum cTLI")

shapiro.test(baseline_clinicVar$Serum_cTLI)
#major violoation of normality
```

Major violation of normality for cTLI. Is it resolved with log-transformation?

```{r}
hist(log(baseline_clinicVar$Serum_cTLI), main = "Distrubition of Serum log(cTLI)")

shapiro.test(log(baseline_clinicVar$Serum_cTLI))

```

Log transformation does not resolve violation. Use non-parametric Wilcoxon test. 

```{r, results='asis'}
summary(tableby(Group ~ Serum_cTLI,
                data = baseline_clinicVar, 
                numeric.test="wt", 
                numeric.stats=c("mean", "sd", "median", "min", "max", "iqr", "q1q3")))

```

```{r}

tli <- ggplot(baseline_clinicVar, aes(x=Group, y=Serum_cTLI, fill = Group)) + 
  geom_boxplot() +
  stat_compare_means(comparisons = list(c("Healthy", "EPI")), 
                     label = "p.signif", 
                     method = "t.test") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  scale_fill_manual(values = plot_cols) +
  theme_bw() +
  labs(y="Serum cTLI (µg/L)", x = "")

tli
```

Serum cTLI varies significantly between groups and the EPI dogs have lower serum concentrations of cTLI compared with healthy controls. As expected. 

### **Serum Cobalamin**

Dogs receiving cobalamin supplementation at the time of sample collection were removed for this analysis.

```{r}

baseline_clinicVar[!baseline_clinicVar$Cobalamin_sup == "Yes", "Serum_Cobalamin"] %>% 
  hist(main = "Distribution of Serum Cobalamin")

baseline_clinicVar[!baseline_clinicVar$Cobalamin_sup == "Yes", "Serum_Cobalamin"] %>% 
  shapiro.test()

```

Though SW test is not violated, the histogram appears skewed, use wt. 

```{r, results='asis'}

summary(tableby(Group ~ Serum_Cobalamin,
                data = baseline_clinicVar[!baseline_clinicVar$Cobalamin_sup == "Yes", ],
                numeric.test="wt",
                numeric.stats=c("mean", "sd", "median", "min", "max", "iqr", "q1q3")))

```

Serum cobalamin varies significantly by group and is lower in the dogs with EPI compared with healthy controls. 

```{r}
b12 <- ggplot(baseline_clinicVar[!baseline_clinicVar$Cobalamin_sup == "Yes", ], aes(x=Group, y=Serum_Cobalamin, fill = Group)) + 
  geom_boxplot() +
  stat_compare_means(comparisons = list(c("Healthy", "EPI")), 
                     label = "p.signif", 
                     method = "t.test") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  scale_fill_manual(values = plot_cols) +
  theme_bw() +
  labs(y="Serum Cobalamin (ng/L)", x="")

b12
  
```

### **Serum Folate**

Dogs receiving folate supplementation at the time of sample collection were removed for this analysis.

```{r}
baseline_clinicVar[!baseline_clinicVar$Folate_sup == "Yes", "Serum_Folate"] %>% 
  hist(main = "Distribution of Serum Folate")

baseline_clinicVar[!baseline_clinicVar$Folate_sup == "Yes", "Serum_Folate"] %>% 
  shapiro.test()

```

The distribution of serum folate is approximately normal. 

```{r, results='asis'}
summary(tableby(Group ~ Serum_Folate,
                data = baseline_clinicVar[!baseline_clinicVar$Folate_sup == "Yes", ], 
                numeric.test="anova",
                numeric.stats=c("mean", "sd", "median", "min", "max")))

```

There are no significant differences in serum folate concentrations between dogs with EPI and healthy controls.

```{r}
folate <- ggplot(baseline_clinicVar[!baseline_clinicVar$Folate_sup == "Yes", ], 
                 aes(x=Group, y=Serum_Folate, fill = Group)) + 
  geom_boxplot() +
  stat_compare_means(comparisons = list(c("Healthy", "EPI")), 
                     label = "p.signif", 
                     method = "t.test") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  scale_fill_manual(values = plot_cols) +
  theme_bw() +
  scale_fill_manual(values = plot_cols) +
  labs(y="Serum Folate (µg/L)", x="") +
  ggtitle("Serum Folate")

folate
  
```

# **UNSUPERVISED ANALYSIS**

All metabolite concentrations were log-transformed prior to analysis.

```{r}
metab_log <- log(metab)

metab_log <- data.frame(t(metab_log))
```

## **PRINCIPAL COMPONENT ANALYSIS**


```{r, include=F}

#prepare data
metab_pca <- metab_log

metab_pca$Group <- EPI_groups$Group[match(rownames(metab_pca), rownames(EPI_groups))]
table(metab_pca$Group)
metab_pca$Group <- factor(metab_pca$Group, levels = c("Healthy", "EPI"))

```

```{r}
#perform PCA
pca <- PCA(metab_log[, -760], graph = F, scale.unit = T)

fviz_eig(pca)

summary(pca)

```

```{r}

pc <- data.frame(cbind(PC1=pca$ind$coord[, 1], PC2=pca$ind$coord[, 2]))
pc$Group <- EPI_groups$Group[match(rownames(pc), EPI_groups$Sample_ID)]

pca_plot <- ggplot(data = pc, aes(x = PC1, y = PC2, color = Group)) +
  scale_color_manual(values = plot_cols) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  guides(color = guide_legend(title = "Group")) +
  scale_shape_manual(values = c(15, 16, 16, 17, 18)) +
  geom_point(alpha = 1, size = 3) + 
  xlab("PC 1 (11.5%)") + 
  ylab("PC 2 (9.9%)") +
  theme_minimal() +
  theme_bw() + 
  theme(legend.title=element_text(size=14), 
    legend.text=element_text(size=12))

```


## **HEATMAP WITH CLUSTERING BY EUCLIDIAN DISTANCES**

```{r}

metab_heat <- metab_log

metab_heat$Group <- EPI_groups$Group[match(rownames(metab_heat), rownames(EPI_groups))]

annotation_colors <- list(Group = c(Healthy = "royalblue", EPI = "orange"))

annotation_col <- as.data.frame(metab_heat$Group)

names(annotation_col)[1] <- "Group"

rownames(annotation_col) <- rownames(metab_heat)

metab_heat <- metab_heat[, -760]
metab_heat <- data.frame(t(metab_heat))

ph <- pheatmap(
  mat               = metab_heat,
  color             = inferno(10),
  border_color      = NA,
  show_colnames     = T,
  annotation_col    = annotation_col,
  annotation_colors = annotation_colors,
  cluster_cols = T,
  cluster_rows = T,
  drop_levels       = TRUE,
  scale             = "row",
  fontsize          = 8,
  annotation_legend = F, 
  fontsize_row = 5,
  fontsize_col = 5,
  show_rownames = F,
  treeheight_row = 0,

)



```
# **DIFFERENTIAL ANALYSIS OF RELATIVE METABOLITE CONCENTRATIONS**

## **T-TESTS**

Welch's t-test will be performed on each serum metabolite to identify those that vary significantly betweem dogs with EPI and healthy controls. The log2-fold difference in means for each metabolite will be caclulated to estimate the effect size. FDR-adjusted p-values (q-values) will be calculated using the BH method.  

```{r, include=F}

metab_log$Group <- EPI_groups$Group[match(rownames(metab_log), rownames(EPI_groups))]


t_test.res <- lapply(metab_log[, -760],
       function(x)
         t.test(x ~ metab_log$Group, paired = F)
)

pval <- sapply(t_test.res, '[[', 'p.value') %>%
  as.data.frame(row.names = names(metab_log[, -760]))

names(pval)[1] <- paste("P-value")

#calculate FDR-corrected p-value
padj <- t(pval) %>%
  p.adjust(method = "BH", n = nrow(pval)) %>%
  as.data.frame(row.names = names(metab_log[, -760]))

names(padj)[1] <- paste("pval_adj_BH")

p_table_t <- cbind(pval, padj)

sig_table_t <- na.omit(p_table_t[p_table_t$`P-value` < 0.05, ])

sig_table_t$Biochemical <- annot$BIOCHEMICAL[match(rownames(sig_table_t), rownames(annot))]

```

How many metabolites varied significantly (P<0.05)?

```{r}

paste(nrow(sig_table_t), "metabolites varied significantly (P<0.05) between dogs with EPI and healthy dogs", sep = " ")
```

How many metabolites had a low FDR (FDR<0.2)?

```{r}

paste(nrow(p_table_t[p_table_t$pval_adj_BH <0.2, ]), 
      "significantly variable metabolites had a FDR<0.2")

```

## **log2 Fold-Changes**

```{r, include=F}

metab_means <- metab %>% t() %>% data.frame() %>%
  mutate_if(is.numeric, log2) %>%
  mutate(Group=EPI_groups$Group) %>% 
  group_by(Group) %>%
   dplyr::summarise(across(
    .cols = is.numeric, 
    .fns = list(Mean = mean), 
    na.rm = T, 
    .names = "{col}_{fn}"
    ))

metab_means <- metab_means[c(2:1), ]

names <- colnames(metab_means)[2:760]
nNames <- length(names)

foldchange.res <- NULL

for (i in names) {
  # print status
  #print(paste("Running entity:", i, "which is", which(names==i), "out of", nNames))
  name <- i
   
  fc1 <- data.frame(metab_means[1,i] - metab_means[2,i])
  fc1$contrast <- "EPI - Healthy"
  names(fc1)[1] <- "Fold-Change"
  fc1$met <- i
  
  foldchange.res <- rbind(foldchange.res, fc1)

}

foldchange.res$met <- gsub("_Mean", "", foldchange.res$met)

```

## **INDENTIFY SIGNIFICANTLY VARIABLE METABOLITES**

```{r, include=F}

p_table_t$met <- rownames(p_table_t)
combined.res <- cbind(p_table_t, foldchange.res) 

combined.res <- combined.res[, c(3, 5, 1:2, 4)]

combined.res$Biochemical <- annot$BIOCHEMICAL[match(combined.res$met, 
                                                     rownames(annot))]

combined.res$SuperPathway <- annot$SUPER.PATHWAY[match(combined.res$met,
                                                       rownames(annot))]

combined.res$SubPathway <- annot$SUB.PATHWAY[match(combined.res$met, 
                                                    rownames(annot))]

combined.res <- combined.res[, c(1, 6:8, 2:5)]

names(combined.res)[6:8] <- c("pval", "FDR", "log2FC")
```


```{r}
sig <- combined.res[combined.res$FDR <0.2, ]
sig <- sig[order(sig$SuperPathway, sig$log2FC), ]

sig$log2FC <- round(sig$log2FC, 2)
sig$pval <- round(sig$pval, 3)
sig$FDR <- round(sig$FDR, 3)

sig_inc <- sig[sig$log2FC > 0, ]
rownames(sig_inc) <- c()

sig_dec <- sig[sig$log2FC < 0, ]
rownames(sig_dec) <- c()

inc <- as.data.frame(table(sig_inc$SuperPathway))
names(inc)[1] <- "SuperPathway"
names(inc)[2] <- "Increased Metabolites"

paste(sum(inc$`Increased Metabolites`),
      "significantly variable metabolites were INCREASED in the sera of dogs with EPI")



dec <- as.data.frame(table(sig_dec$SuperPathway))
names(dec)[1] <- "SuperPathway"
names(dec)[2] <- "Decreased Metabolites"

paste(sum(dec$`Decreased Metabolites`),
      "significantly variable metabolites were DECREASED in the sera of dogs with EPI")


```

```{r}

inc2 <- inc
names(inc2)[2] <- "Significant_Changes"
inc2$color <- "Increased"
inc2 <- inc2[!inc2$Significant_Changes == 0, ]

dec2 <- dec
names(dec2)[2] <- "Significant_Changes"
dec2$color <- "Decreased"
dec2 <- dec2[!dec2$Significant_Changes == 0, ]

freq_tbl <- rbind(inc2, dec2)

freq_tbl$color <- factor(freq_tbl$color, 
                           levels = c("Increased", "Decreased"))

freq_tbl <-freq_tbl[order(-freq_tbl$Significant_Changes), ]

super_path.plot <- ggplot(data = freq_tbl, aes(y = Significant_Changes, 
                                               x = reorder(SuperPathway, -Significant_Changes, sum), 
                                               fill = color)) + 
  geom_bar(stat = "identity") + 
  labs(fill = "") +
  xlab("") +
  ylab("Number of Significant Metabolites") + 
  scale_fill_brewer(palette="Set1") +
  theme_bw() +
  theme(legend.position = "right") + theme(legend.title.align=0.5) +
  theme(axis.text.x=element_text(angle=45,hjust=1))

super_path.plot
```

Plot all statistically significant (FDR<0.2) serum metabolites

```{r, include=F}

sig_metab <- metab_log

keep <- c(sig$met, "Group")

sig_metab <- sig_metab[, keep]

names(sig_metab)[1:114] <- annot$BIOCHEMICAL[match(names(sig_metab[1:114]),
                                                   rownames(annot))] 

base_plot2 <-sig_metab %>%
  gather(Measure, Value, -Group) %>%
  ggplot(aes(x = factor(Group), y = Value, fill = Group)) +
  geom_boxplot() +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = plot_cols) +
  facet_wrap(~Measure, 
             scales = "free") +
  xlab("Group") +
  ylab("Relative Metabolite Concentration (Log, Median-Scaled))") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  theme(strip.text.x = element_text(size = 5))


pdf("S_Fig_sigBoxplots.pdf")
facet_multiple(base_plot2, "Measure",
               ncol = 3,
               scales = "free")
dev.off()


```

Table of all significant metabolites with FDR<0.2 and a large effect size (|log2FC|>0.06; >1.5 fold difference in means)

```{r}

sig <- sig %>% arrange(SuperPathway, SubPathway, log2FC, FDR)

kable(sig[sig$FDR<0.2 & abs(sig$log2FC)>0.6,
          c(2:4, 6:8)], row.names = F)

```


```{r}
#Output results

names(combined.res)[7] <- "qval"

##all t.test results
write.csv(combined.res, "K9EPI_serumMetab_ttest_res.csv")

#significant results (p<0.05)
write.csv(sig, "K9EPI_serumMetab_ttest_sigres.csv")
```


# **METABOLITE SET ENRICHMENT ANALYSIS**


Adapted from gene set enrichment analysis using the gsea package. 

 * Rank metabolites by the product of the log2FC and log(p-val) from the t-test
 * The top end of the list are highly significant AND have high positive fold change
 * The bottom end of the list are highly significant AND have negative fold change
 * gsea function used to detect significant enrichement of metabolic sub-pathways among significant results
 * calculated normalized enrichment score 

```{r, include=F}

diff_enrich <- combined.res

diff_enrich$rank <- -log(diff_enrich$pval)*diff_enrich$log2FC

diff_enrich <- diff_enrich[order(diff_enrich$rank, decreasing = T), ]

#create ranking vector from product of log2FC*pval

rank <- diff_enrich[, 9]
names(rank) <- diff_enrich$Biochemical
# head(rank)
# tail(rank)

#Now generate lists of metabolites in each metabolic sub-pathway

pathways <- diff_enrich[, 2:4]

pathways <- pathways[order(pathways$SubPathway, pathways$Biochemical),]

#pathways$Biochemical <- as.character(pathways$Biochemical)
pathways$SubPathway <- as.character(pathways$SubPathway)

pathways <- split(pathways$Biochemical, as.character(pathways$SubPathway))

class(pathways)

set.seed(123)
fgseaRes <- fgsea(pathways = pathways, 
                  stats = rank,
                  minSize=2,
                  maxSize=20)


#saveRDS(fgseaRes, "K9EPI_serumMetab_pathways_res.RDS")

```

```{r}
fgseaRes_out <- data.frame(fgseaRes)

names(fgseaRes_out)[3] <- "qval"

fgseaRes_out <- apply(fgseaRes_out,2,as.character)

write.csv(fgseaRes_out, file = "fgseaRes.csv")
```

Plot normalized pathway enrichment scores

```{r}

topPathwaysUp <- fgseaRes[ES > 0][head(order(padj), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(padj), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))


topPathways <- unlist(topPathways)

fgseaRes_top20 <- fgseaRes[fgseaRes$pathway %in% topPathways, ]

fgseaRes_top20 <- fgseaRes_top20[order(fgseaRes_top20$NES), ]

fgseaRes_top20$pathway <- as.factor(fgseaRes_top20$pathway)

fgseaRes_top20$pathway <- factor(fgseaRes_top20$pathway, 
                                 levels = fgseaRes_top20$pathway[order(fgseaRes_top20$NES)])

gsea_plot <- ggplot(data = fgseaRes_top20, aes(y = fgseaRes_top20$NES, 
                                               x = fgseaRes_top20$pathway, 
                                               fill  = ifelse(fgseaRes_top20$padj 
                                                             <0.2, 
                                                             "FDR q<0.2",
                                                             "FDR q≥0.2"))) +
  geom_bar(stat = "identity", position = "dodge") + ylim(-3, 3) + 
  ylim(-2, 2) +
  coord_flip() + 
  labs(fill = "") +
  xlab("") +
  ylab("Normalized Enrichment Score") +
  scale_fill_manual(values=c("orange", "grey")) +
  theme_bw() +
  theme(legend.position = "bottom") + theme(legend.title.align=0.5)
      
gsea_plot


```



